Are environmental and geographic variables effective surrogates for genetic variation in conservation planning?

### 2016-03-18 article/article.Rmd
+ @Zongle comments incorperated to article + final dataset added to ms

# set options
options(error=NULL)
```

@@

## Introduction
<!--  main text -->
The act of setting aside areas for preservation spearheads modern conservation efforts [@r440]. They buffer species from anthropogenic impacts and provide places for populations of species to persist. Intra-specific genetic variation underpins species persistence, and there has been increasing interest in understanding how to design protected areas networks in a way that maximizes this variation [@r67; @r3; @r61]. Broadly speaking, genetic variation can be classified as either adaptive or neutral [reviewed in @r48]. Adaptive genetic variation is associated with loci that affect fitness. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that do not affect fitness. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can secure different genetic lineages and avoid the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. However, genetic data requires considerable investment to obtain and analyze. As a consequence, conservation planners do not typically use genetic data to inform prioritizations [@r61]. Thus prioritizations may fail to capture species' intra-specific genetic variation, and ultimately fail to deliver long-term conservation outcomes.

Recently, conservation scientists have been investigating surrogates for genetic variation to generate better prioritizations without the overheads associated with genetic data. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been used to guide reserve selection [eg. habitat conditions; @r247; @r2]. However, their effectiveness as surrogates in conservation planning remains unverified. On the other hand, neutral genetic variation arises from a reduction in gene flow between individuals. Surrogates for neutral variation are based on variables that predict the level of connectivity between planning units. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have less connectivity, exchange less genetic material, and in turn have less genetic variation in common. Based on this theory, geographic distance-based surrogates have been found to improve prioritizations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain; eg. @r474]--the systems commonly featured in real-world conservation planning exercises. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, conservation planners may waste precious resources on places that do little to benefit biodiversity.

Here we investigate the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation in the context of conservation planning. First, we examine the relationship between surrogates and gene variation, and verify that these variables can be used as surrogates. Second, we determine if the inclusion of surrogate-based targets in a variety of conservation planning scenarios with increasingly levels of realism can result in prioritizations that secure more genetic variation. If these surrogates can be shown to improve the intra-specific genetic variation captured in a prioritization, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

## Methods
### Study system
We used species distribution and genomic data collected by the IntraBioDiv project in the European Alps [@r451; see @r452 and @r478 for methodological details]. This data set was particularly well suited for our study, because it provides spatially explicit genomic data for multiple species with vastly different demographic characteristics and evolutionary histories across a broad geographic area with large environmental gradients. As part of this project, data was collected for 27 alpine plant species using a 20' longitude by 21' latitude grid (approx. 22.3 km $\times$ 25 km; Figure S1). They visited every second grid cell, and if a species was detected in a cell, samples were collected from three individuals. They genotyped samples using amplified fragment length polymorphisms [AFLP; @r453], and constructed matrices denoting the presence/absence of polymorphisms at loci for each species. Thus for each second grid cell, for each of the species present in the grid cell, the data set contains information describing the genomic variation of individuals in the cell. We used these grid cells as planning units to develop prioritizations. To reduce computational effort, we only used grid cells that contained at least a single individual ($n = `r nrow(grid.DF)`$). The majority of our analyses were performed using the \texttt{R} software package [version `r paste0(version$major, '.', version$minor)`; @r192].

### Landscape data
We examined the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation (respectively). To describe the geographic location of each grid cell, we projected the grid into an equi-distant coordinate system ($\texttt{ESRI:102031}$), calculated their centroids, and extracted their two-dimensional coordinates. To describe the environmental characteristics of each grid cell, we obtained 19 bioclimatic variables [approx. 1 km resolution; @r33]. These layers and the grid cells were projected into an equal-area coordinate system ($\texttt{ESRI:102014}$). The bioclimatic variables were clipped to the extent of the grid and subject to a principal components analysis (PCA; using \texttt{ArcMap} version 10.2.2; Table S1). The first two principal components (PCs) cumulatively explained `r round(pca.DF[[4]][2],1)` % of the variation and were used to generate two new layers. We calculated the average value for each PC layer in each grid cell, and used these values to characterize the main themes of climatic variation among the grid cells (Figure S2). To examine if the influence of opportunity cost, we calculated the total human population density inside each planning unit [obtained at 1 km resolution from the Global Rural-Urban Mapping Project; \texttt{GRUMP V1}; @r472].

### Outlier locus detection
To investigate the effectiveness of the surrogates for adaptive and neutral genetic variation, we first needed to identify which of the sampled loci--if any--were under selection. We used outlier locus detection methods [compared in @r454; reviewed in @r480] to achieve this. The basic premise underpinning these analyses is that neutral loci are expected to exhibit a specific level of variation among populations ($F_{st}$), and loci that deviate from this expectation should be under selection. The advantage of these methods--in contrast with environmental association analyses [eg. @r475]--is that they do not use environmental data to identify loci under selection, which would have biased our analysis. In an earlier study, Bothwell _et al._ [-@r455] applied such outlier detection methods to one of the species in this data set (_Gentiana nivalis_). Here, we adopted a similar methodology and applied it to each of the 27 species in the data set.

First, we grouped individuals into genetic lineages for each species [using a methodology similar to that implemented in \texttt{prabclust R} package; @r479]. The purpose of this step was to avoid falsely classifying loci as adaptive. For each species, we ordinated the genomic data and fit models to the ordinations to fit to turn group individuals into populations. Specifically, we used non-metric multi-dimensional scaling [NMDS; implemented in the $\texttt{vegan R}$ package; @r458] using Gower distances [@r464; using the $\texttt{cluster R}$ package to accommodate sparsity; @r457] to transform the binary loci data into continuous variables (Table S2). To ensure that the ordinations described a sufficient amount of the genomic variation, we ran successive scaling analyses with increasing dimensionality until this was achieved (maximum stress value $\leq$ `r nmds.params.LST[[MODE]]$max.stress`; `r nmds.params.LST[[MODE]]$trymax` random starts for each analysis). We then fit a suite of Gaussian mixture to the ordinations to estimate the probability that individuals belonged to different populations [assuming different covariance structures implemented in the \texttt{mclust} R package; @r477; @r476]. For each parametrization, we fit a set of models assuming varying numbers of total populations (\textit{G}= 1--`r max(mclust.params.LST[[MODE]]$G)`). We identified the most parsimonious model for each species (lowest Bayesian Information Criterion $BIC$ statistic; Figures S4), and used them to group individuals to populations (Figure S5).

We identified the loci under selection (Table S3). Since outlier locus detection methods compare allele frequencies between populations, we were unable to examine species that contained a single population. For these species ($n = `r sum(sapply(spp.Mclust.LST, function(x) return(ncol(x$mclust$z)==1)))`$), we assumed that all loci were neutral. For species with more than a single population, we ran $\texttt{Bayescan}$ [version 2.1; 1:1 prior odds; `r bayescan.params.LST[[MODE]]$nbp` pilot runs; `r bayescan.params.LST[[MODE]]$burn` burn-in iterations; `r bayescan.params.LST[[MODE]]$n` post-burn-in iterations thinned by `r bayescan.params.LST[[MODE]]$thin` iterations; @r456]. To avoid false-positives, we omitted loci where the global frequency of the minor allele was $\geq 0.05$ (following guidelines in the user manual) and assumed they were neutral. Additionally, we omitted individuals from this analysis when their population membership was uncertain (maximum membership probability $\l `r mclust.params.LST[[MODE]]$probability.threshold`$). We ran `r bayescan.params.LST[[MODE]]$reps` replicates per species using a suitable false discovery rate to ensure convergence [FDR $\leq `r bayescan.params.LST[[MODE]]$fdr`$; @r466].

@@

To determine if the use of surrogate-based targets can actually improve prioritizations in real-world conservation planning scenarios, we investigated the effectiveness of surrogate-based targets using three scenarios with increasing levels of complexity. To assess the benefits of surrogate-based targets under the simplest of circumstances, our first scenario involved generating single-species prioritizations for each species. However, most large-scale prioritizations involve a comprehensive set of species. This property may lead to prioritizations that secure a greater proportion of intra-specific genetic variation for each species by returning solutions that preserve the same species in multiple communities. To test for this effect, our second scenario involved generating multi-species prioritizations. Finally, most large-scale prioritizations also explicitly consider opportunity cost to ensure prioritizations are cost-effective. Our third scenario involved generating multi-species prioritizations that satisfied targets for minimal opportunity cost.

For each scenario, we generated three types of prioritizations. We generated prioritizations using only amount-based targets, both amount- and surrogate-based targets, and both amount- and genetic-based targets. The purely amount-based prioritizations serve as our baseline--they represent prioritizations typically generated in conservation planning exercises. We used `r (rapr.params.LST[[MODE]][['single.species']]$amount.target*100)` % amount-based targets in all prioritizations to ensure an adequate proportion of habitat was secured for each species. We set surrogate- and genetic-based targets as `r (rapr.params.LST[[MODE]][['single.species']]$amount.target*100)` % to ensure that most of the respective variation was secured for each species. To accomodate variation in the high number of optimal solutions for the single-species amount-based prioritizations, we generated `r (rapr.params.LST[[MODE]][['single.species']]$amount.replicates)` replicate prioritizations. Since all other contexts were likely to be associated with few optimal solutions, we generated only `r rapr.params.LST[[MODE]][['multi.species']]$genetic.replicates` replicates.

## Statistical analysis
We investigated the effectiveness of prioritizations generated using surrogate-based targets using generalized linear mixed-effects models [GLMMs; using a logit link and the $\texttt{lme4}$ R package; @r469]. To determine  if the environmental and geographic variables were effective surrogates, we fit a full model to the proportion of genetic variation that the first collection of prioritizations secured using categorical variables to indicate the surrogate-based targets used to generate them, the type of genetic variation (adaptive or neutral), the method used to generate the prioritization (optimal or random), and the interactions between these variables. To determine if surrogate-based targets improved the effectiveness of prioritizations under various scenarios, we fit a full model to the proportion of genetic variation that second collection of prioritizations secured using categorical variables to indicate the scenario (single-species, multi-species, multi-species with cost), the combination of targets used to generate the prioritizations (amount-based targets, amount and surrogate-based targets, amount and genetic-based targets), the type of genetic variation measured (adaptive or neutral), and the interactions between them. All model contained intercept-level random effects to accommodate variation between species. Both full models were subject to a step-wise term deletion routine using $\chi^{2}$ tests to assess significance (terms were retained when $P < 0.05$). Finally, we conducted post-hoc analysis on each full model using Tukey contrasts with a Bonferroni correction [using the $\texttt{multcomp}$ R package; @r470].

## Results
<!-- Surrogacy analysis -->
Prioritizations generated using greater environmental and geographic surrogate-based targets were found to secure greater proportions of adaptive and neutral genetic variation (Figure 1). The proportion of genetic variation that these prioritizations secured was substantially more than the genetic variation secured by randomly generated prioritizations (${\chi^{2}}_{`r NA`,`r NA`} = `r  NA`$, $P < `r NA`$). Generally, prioritizations needed to capture at least `r NA` % of the environmental and geographic surrogates to secure 90 % of the species' genetic variation. Additionally, we detected differences in the effectiveness of the environmental and geographic surrogates (${\chi^{2}}_{`r NA`,`r NA`} = `r  NA`$, $P < `r NA`$). Increases in environmental surrogate-based targets resulted prioritizations that secured more adaptive genetic variation, than geographic surrogate-based targets for neutral variation. These results demonstrate that environmental and geographic variables are effective surrogates for genetic variation for these plant species.

>>>>

# set options
options(error=NULL)

```

@@

## Introduction
<!--  main text -->
The act of setting aside areas for preservation spearheads modern conservation efforts [@r440]. Protected area networks buffer species from anthropogenic impacts and provide places for populations of species to persist. Intra-specific genetic variation underpins species persistence, and there has been increasing interest in understanding how to design protected area networks in a way that maximizes this variation [@r67; @r3; @r61]. Broadly speaking, genetic variation can be classified as either adaptive or neutral [reviewed in @r48]. Adaptive genetic variation is associated with loci that affect fitness. By preserving existing patterns of adaptive variation, protected areas can ensure that populations with particularly beneficial adaptations are not lost [@r3]. In contrast, neutral genetic variation is associated with loci that do not affect fitness. Neutral variation often reflects the evolutionary history of different populations. By preserving existing patterns of neutral variation, protected areas can secure different genetic lineages and avoid the adverse effects of low genetic diversity [eg. inbreeding depression; @r67]. However, genetic data requires considerable investment to obtain and analyze. As a consequence, conservation planners do not typically use genetic data to inform prioritizations [@r61]. Thus prioritizations may fail to capture species' intra-specific genetic variation, and ultimately fail to deliver long-term conservation outcomes.

Recently, conservation scientists have been investigating surrogates for genetic variation to generate better prioritizations without the overheads associated with genetic data. Since adaptive genetic variation is ultimately driven by extrinsic selection pressures, environmental variables have been used to guide reserve selection [eg. habitat conditions; @r247; @r2]. However, their effectiveness as surrogates in conservation planning remains unverified. On the other hand, neutral genetic variation arises from a reduction in gene flow between individuals. Surrogates for neutral variation are based on variables that predict the level of connectivity between planning units. For instance, the isolation by distance theory [IBD; @r259] predicts that populations further apart will have less connectivity, exchange less genetic material, and in turn have less genetic variation in common. Based on this theory, geographic distance-based surrogates have been found to improve prioritizations in an island network [@r200]. However this remains untested in spatially contiguous systems where connectivity is complicated by additional factors [eg. terrain; eg. @r474]--the very systems commonly featured in real-world conservation planning exercises [eg. @r197]. Before conservation planners can reliably use surrogates for genetic data in conservation planning, the effectiveness of these surrogates needs to be verified. Otherwise, conservation planners may waste precious resources on protecting places that provide little benefit to biodiversity.

Here we investigate the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation in the context of conservation planning. First, we examine the relationship between surrogates and genetic variation, and verify whether these variables are effective surrogates. Second, we determine if the inclusion of surrogates in various conservation planning scenarios can improve the proportion of genetic variation they secure. If these surrogates can improve the intra-specific genetic variation captured in a prioritization, then conservation planners may use these surrogates when genetic data is not available, and in turn deliver more effective protected areas.

## Methods
### Study system
We used species distribution and genomic data collected by the IntraBioDiv project in the European Alps [@r451; see @r452 and @r478 for a detailed explanation of data collection methods]. This data set was particularly well suited for our study because it contains genetic data for a diverse range of species that were collected using the same sampling regime. Additionally, the study area covers a broad geographic domain that encompasses  attitudinal gradients which provide substantial variation to examine the surrogates. As part of this project, data was collected for 27 alpine plant species using a 20' longitude by 21' latitude grid (approx. 22.3 km $\times$ 25 km; Figure S1). Project members visited every second grid cell, and if a species was detected in a cell, samples were collected from three individuals. They genotyped samples using amplified fragment length polymorphisms [AFLP; @r453], and constructed matrices denoting the presence/absence of polymorphisms at loci for each species. Thus the data set contains information describing the genomic variation of individuals for each species present in every second grid cell. We used these grid cells as planning units to develop prioritizations. To reduce computational effort, we used only grid cells that contained at least a single individual ($n = `r nrow(grid.DF)`$). The majority of our analyses were performed using the \texttt{R} software package [version `r paste0(version$major, '.', version$minor)`; @r192].

### Landscape data
We examined the effectiveness of environmental and geographic surrogates for adaptive and neutral genetic variation (respectively). To describe the geographic location of each grid cell, we projected the grid into an equi-distant coordinate system ($\texttt{ESRI:102031}$), calculated their centroids, and extracted their two-dimensional coordinates. To describe the environmental characteristics of each grid cell, we obtained 19 bioclimatic variables [approx. 1 km resolution; @r33]. These layers and the grid cells were projected into an equal-area coordinate system ($\texttt{ESRI:102014}$). The bioclimatic variables were clipped to the extent of the grid and subject to a principal components analysis (PCA; using \texttt{ArcMap} version 10.2.2; Table S1). The first two principal components (PCs) cumulatively explained `r round(pca.DF[[4]][2],1)` % of the variation and were used to generate two new layers. We calculated the average value for each PC layer in each grid cell, and used these values to characterize the main themes of climatic variation among the grid cells (Figure S2). The cost of purchasing land for preservation often exerts a strong influence on which planning units are recommended for prioritization [reviewed in @r482]. To examine the potential effects of cost, we calculated the total human population density inside each planning unit [obtained at 1 km resolution from the Global Rural-Urban Mapping Project; \texttt{GRUMP V1}; @r472] and used this to represent acquisition cost.

### Adaptive and neutral genetic data
To investigate the effectiveness of the surrogates for adaptive and neutral genetic variation, we first needed to identify which of the sampled loci--if any--were under selection. We used outlier detection methods [compared in @r454; reviewed in @r480] to achieve this. The basic premise underpinning these analyses is that neutral loci are expected to exhibit a specific level of variation among populations ($F_{st}$), and loci that deviate from this expectation should be under selection. The advantage of these methods--in contrast with environmental association analyses [eg. @r475]--is that they do not use environmental data to identify loci under selection, which would have biased our analysis. In an earlier study, Bothwell _et al._ [-@r455] applied such outlier detection methods to one of the species in this data set (_Gentiana nivalis_). Here, we adopted a similar methodology and applied it to each of the 27 species in the data set.

First, we grouped individuals into genetic lineages for each species [using a methodology similar to that implemented in \texttt{prabclust R} package; @r479]. The purpose of this step was to avoid falsely classifying loci as adaptive. For each species, we ordinated the genomic data, and then fit models to the ordinated data to group individuals into populations. Specifically, we used non-metric multi-dimensional scaling [NMDS; implemented in the $\texttt{vegan R}$ package; @r458] using Gower distances [@r464; using the $\texttt{cluster R}$ package to accommodate sparsity; @r457] to transform the binary loci data into continuous variables (Table S2). To ensure that the ordinations described a sufficient amount of the genomic variation, we ran successive scaling analyses with increasing dimensionality until this was achieved (maximum stress value $\leq$ `r nmds.params.LST[[MODE]]$max.stress`; `r nmds.params.LST[[MODE]]$trymax` random starts for each analysis). We then fit a suite of Gaussian mixture to the ordinations to estimate the probability that individuals belonged to different populations [assuming different covariance structures implemented in the \texttt{mclust} R package; @r477; @r476]. For each parametrization, we fit a set of models assuming varying numbers of total populations (\textit{G}= 1--`r max(mclust.params.LST[[MODE]]$G)`). We identified the most parsimonious model for each species (lowest Bayesian Information Criterion $BIC$ statistic; Figures S4), and used them to group individuals to populations (Figure S5).

After grouping individuals into populations, we identified the loci under selection (Table S3). Since outlier detection methods compare allele frequencies between populations, we were unable to examine species that contained a single population. For these species ($n = `r sum(sapply(spp.Mclust.LST, function(x) return(ncol(x$mclust$z)==1)))`$), we assumed that all loci were neutral. For species with more than a single population, we ran $\texttt{Bayescan}$ [version 2.1; 1:1 prior odds; `r bayescan.params.LST[[MODE]]$nbp` pilot runs; `r bayescan.params.LST[[MODE]]$burn` burn-in iterations; `r bayescan.params.LST[[MODE]]$n` post-burn-in iterations thinned by `r bayescan.params.LST[[MODE]]$thin` iterations; @r456]. To avoid false-positives, we omitted loci where the global frequency of the minor allele was $\geq 0.05$ (following guidelines in the user manual) and assumed they were neutral. Additionally, we omitted individuals from this analysis when their population membership was uncertain (maximum membership probability $< `r mclust.params.LST[[MODE]]$probability.threshold`$). We ran `r bayescan.params.LST[[MODE]]$reps` replicates per species using a suitable false discovery rate to ensure convergence [FDR $\leq `r bayescan.params.LST[[MODE]]$fdr`$; @r466].

@@

To determine if the use of surrogate-based targets can actually improve prioritizations in real-world conservation planning scenarios, we investigated the effectiveness of surrogate-based targets using three scenarios with increasing levels of complexity. To assess the benefits of surrogate-based targets under the simplest of circumstances, our first scenario involved generating single-species prioritizations for each species. However, most large-scale prioritizations involve a comprehensive set of species. By forcing solutions to secure individuals in multiple communities, this may result in prioritizations that secure a greater proportion of intra-specific genetic variation for each species. To test for this effect, our second scenario involved generating multi-species prioritizations. Finally, most large-scale prioritizations also explicitly consider opportunity cost to ensure prioritizations are cost-effective. Our third scenario involved generating multi-species prioritizations that satisfied targets for minimal opportunity cost.

For each scenario, we generated three types of prioritizations using (1) only amount-based targets, (2) both amount- and surrogate-based targets, and (3) both amount- and genetic-based targets. The purely amount-based prioritizations served as our baseline--representing prioritizations typically generated in conservation planning exercises. We used `r (rapr.params.LST[[MODE]][['single.species']]$amount.target*100)` % amount-based targets in all prioritizations to ensure an adequate proportion of habitat was secured for each species. We set surrogate- and genetic-based targets as `r (rapr.params.LST[[MODE]][['single.species']]$amount.target*100)` % to ensure that most of the respective variation was secured for each species. To accomodate variation in the high number of optimal solutions for the single-species amount-based prioritizations, we generated `r (rapr.params.LST[[MODE]][['single.species']]$amount.replicates)` replicate prioritizations. Since all other contexts were likely to be associated with few optimal solutions, we generated only `r rapr.params.LST[[MODE]][['multi.species']]$genetic.replicates` replicates.

## Statistical analysis
We investigated the effectiveness of prioritizations generated using surrogate-based targets using generalized linear mixed-effects models [GLMMs; using a logit link and the $\texttt{lme4}$ R package; @r469]. To determine  if the environmental and geographic variables were effective surrogates, we fit a full model to the proportion of genetic variation that the first collection of prioritizations. This full model contained categorical predictor variables to indicate the surrogate-based targets used to generate them, the type of genetic variation (adaptive or neutral), the method used to generate the prioritization (optimal or random), and the interactions between these variables. To determine if surrogate-based targets improved the effectiveness of prioritizations under various scenarios, we fit a full model to the proportion of genetic variation that second collection of prioritizations secured. This second full model contained categorical predictor variables to indicate the scenario (single-species, multi-species, multi-species with cost), the combination of targets used to generate the prioritizations (amount-based targets, amount and surrogate-based targets, amount and genetic-based targets), the type of genetic variation measured (adaptive or neutral), and the interactions between them. All model contained intercept-level random effects to accommodate variation between species. Both full models were subject to a step-wise term deletion routine using $\chi^{2}$ tests to assess significance (terms were retained when $P < 0.05$). Finally, we conducted post-hoc analysis on each full model using Tukey contrasts with a Bonferroni correction [using the $\texttt{multcomp}$ R package; @r470].

## Results
<!-- Surrogacy analysis -->
Prioritizations generated using higher environmental and geographic surrogate-based targets were found to secure greater proportions of adaptive and neutral genetic variation (Figure 1). The proportion of genetic variation that these prioritizations secured was substantially more than the genetic variation secured by randomly generated prioritizations (${\chi^{2}}_{`r NA`,`r NA`} = `r  NA`$, $P < `r NA`$). Generally, prioritizations needed to capture at least `r NA` % of the environmental and geographic surrogates to secure 90 % of the species' genetic variation. Additionally, we detected differences in the effectiveness of the environmental and geographic surrogates (${\chi^{2}}_{`r NA`,`r NA`} = `r  NA`$, $P < `r NA`$). Increases in environmental surrogate-based targets resulted prioritizations that secured more adaptive genetic variation, than geographic surrogate-based targets for neutral variation. These results demonstrate that environmental and geographic variables are effective surrogates for genetic variation for these plant species.
